Temporal graph realization from fastest paths

### 2024-03-29 SAND2024/SAND24-cameraReady.tex
Camera ready version for SAND

\author{Hendrik~Molter}{Department of Computer Science, Ben-Gurion~University~of~the~Negev, Beer-Sheva, Israel}{molterh@post.bgu.ac.il}{https://orcid.org/0000-0002-4590-798X}{Supported by the ISF, grant No.~1456/18, and the ERC, grant number 949707.}

@@

\subparagraph{Related work.} Graph realization problems on static graphs have been studied since the 1960s. We provide an overview of the literature in the introduction. 
To the best of our knowledge, we are the first to consider graph realization problems in the temporal setting. However, many other connectivity-related problems have been studied in the temporal setting~\cite{Mertzios-transitivity21,Akrida-explorer-21,enright2021deleting,MolterRZ21,klobas2023interference,deligkas2022optimizing,erlebach2021temporal,Flu+19a,Zsc+19,CasteigtsCS22,FuchsleMNR22}, most of which are much more complex and computationally harder than their non-temporal counterparts, and some of which do not even have a non-temporal counterpart.

@@

\begin{figure}%[t]
\noindent\makebox[\textwidth]{
\centering
\includegraphics{fig-W1-reduction}
    }
    \caption{Illustration of part of the underlying graph $G$ and a possible labeling. Edges incident with vertices $\hat{v}_1,\hat{v}_2$ of connector gadgets are omitted. Gray vertices form a feedback vertex set.
    The double line connections, between a vertex $v_{i-1}^j$ in the verification gadget, and $u_1^3$ in the edge selection gadget, 
    and, between a vertex $u_2^3$ in the edge selection gadget, and $v_{i}^j$ in the verification gadget,
    consist of $5n$ vertices $a_1^{j,i,3},a_2^{j,i,3},\dots,a_{5n}^{j,i,3}$ 
    and $b_1^{j,i,3},b_2^{j,i,3},\dots,b_{5n}^{j,i,3}$, respectively.
    }\label{fig:hardness1}
\end{figure}

>>>>

\author{Hendrik~Molter}{Department of Computer Science, Ben-Gurion~University~of~the~Negev, Beer-Sheva, Israel}{molterh@post.bgu.ac.il}{https://orcid.org/0000-0002-4590-798X}{Supported by the ISF, grant nr.~1456/18, and by the European Union's Horizon Europe research and innovation programme under grant agreement 949707.}

@@

\subparagraph{Related work.} Graph realization problems on static graphs have been studied since the 1960s. We provide an overview of the literature in the introduction. 
To the best of our knowledge, we are the first to consider graph realization problems in the temporal setting. However, many other connectivity-related problems have been studied in the temporal setting~\cite{Mertzios-transitivity21,Akrida-explorer-21,enright2021deleting,MolterRZ21,klobas2023interference,deligkas2022optimizing,erlebach2021temporal,Flu+19a,Zsc+19,CasteigtsCS22,FuchsleMNR22,KMZ23}, most of which are much more complex and computationally harder than their non-temporal counterparts, and some of which do not even have a non-temporal counterpart.

@@

\begin{figure}%[t]
\noindent\makebox[\textwidth]{
\centering
\includegraphics{fig-W1-reduction_nolabels}
    }
    \caption{Illustration of part of the underlying graph $G$ and a possible labeling. Edges incident with vertices $\hat{v}_1,\hat{v}_2$ of connector gadgets are omitted. Gray vertices form a feedback vertex set.
    The double line connections, between a vertex $v_{i-1}^j$ in the verification gadget, and $u_1^3$ in the edge selection gadget, 
    and, between a vertex $u_2^3$ in the edge selection gadget, and $v_{i}^j$ in the verification gadget,
    consist of $5n$ vertices $a_1^{j,i,3},a_2^{j,i,3},\dots,a_{5n}^{j,i,3}$ 
    and $b_1^{j,i,3},b_2^{j,i,3},\dots,b_{5n}^{j,i,3}$, respectively.
    }\label{fig:hardness1}
\end{figure}

### 2024-03-29 SAND2024/SAND24-longversion.tex
Camera ready version for SAND

In this paper we consider \emph{periodic} temporal graphs, \ie temporal graphs in which the temporal availability of each edge of the underlying graph is periodic. 
Many natural and technological systems exhibit a periodic temporal behavior. For example, in railway networks an edge is present at a time step $t$ if and
only if a train is scheduled to run on the respective rail segment at time $t$~\cite{Arrighi2023Multi}. 
Similarly, a satellite, which makes pre-determined periodic movements, can establish a communication link (\ie a temporal edge) with another satellite whenever they are sufficiently close to each other; the existence of these communication links is also periodic. 
In a railway (resp.~satellite) network, a fastest temporal path from $u$ to $v$ represents the fastest railway connection between two stations 
(resp.~the quickest communication delay between two moving satellites). 
Furthermore, periodicity appears also in (the otherwise quite complex) social networks which describe the dynamics of people meeting~\cite{snapnets,sapiezynski2015tracking}, as every person individually follows mostly a daily routine~\cite{Arrighi2023Multi}.

Although periodic temporal graphs have already been studied 
(see~\cite[Class 8]{casteigts2012time} and~\cite{Arrighi2023Multi,ErlebachS20,morawietz2021timecop,morawietz2020timecop}), 
we make here the first attempt to understand the complexity of a graph realization problem in the context of temporal graphs. 
Therefore, we focus in this paper on the most fundamental case, where all edges have the same period $\Delta$ 
(while in the more general case, each edge $e$ in the underlying graph has a period $\Delta_e$).
As it turns out, the periodic temporal graph realization problem with respect to a given $n \times n$ matrix $D$ of the fastest duration times has a very different computational complexity behavior than the classic graph realization problem with respect to shortest path distances in static graphs.

@@

\subparagraph{Further parameterizations.} There are several structural parameters which can be considered to obtain tractability which are either larger or incomparable to the feedback vertex number.
\begin{itemize}
    \item The \emph{vertex cover number} measures the distance to an independent set, on which we trivially only have no-instances of our problem. We believe this is a promising parameter to obtain tractability.
    \item The \emph{tree-depth} measures ``star-likeness'' of a graph and is incomparable to both the feedback vertex number and the feedback edge number. We leave the parameterized complexity of our problem with respect to this parameter open.
    \item Parameters that measure ``path-likeness'' such as the \emph{pathwidth} or the \emph{vertex deletion distance to disjoint paths} are also natural candidates to investigate.
\end{itemize}
Furthermore, we can consider combining a structural parameter with $\Delta$. Our NP-hardness reduction (\cref{thm:NPhardness}) produces instances with constant $\Delta$, so as a single parameter $\Delta$ cannot yield fixed-parameter tractability. However, in our parameterized hardness reduction (\cref{thm:W1wrtFVS}) the value for $\Delta$ in the produced instance is large. This implies that our result does not rule out e.g.\ fixed-parameter tractability for the combination of the treewidth and $\Delta$ as a parameter. We believe that investigating such parameter combinations is a promising future research direction.

\subparagraph{Further problem variants.}
There are many natural variants of our problem that are well-motivated and warrant consideration. In the following, we give two specific examples. We believe that one of the most natural generalizations of our problem is to allow more than one label per edge in every $\Delta$-period. A well-motivated variant (especially from the network design perspective) of our problem would be to consider the entries of the duration matrix $D$ as upper-bounds on the duration of fastest paths rather than exact durations. 
Our work gives a starting point for many interesting future research directions such as the two mentioned examples.

>>>>

In this paper we consider \emph{periodic} temporal graphs, \ie temporal graphs in which the temporal availability of each edge of the underlying graph is periodic. 
Many natural and technological systems exhibit a periodic temporal behavior. For example, in railway networks an edge is present at a time step $t$ if and
only if a train is scheduled to run on the respective rail segment at time $t$~\cite{Arrighi2023Multi}. 
Similarly, a satellite, which makes pre-determined periodic movements, can establish a communication link (\ie a temporal edge) with another satellite whenever they are sufficiently close to each other; the existence of these communication links is also periodic. 
In a railway (resp.~satellite) network, a fastest temporal path from $u$ to $v$ represents the fastest railway connection between two stations 
(resp.~the quickest communication delay between two moving satellites). 
Furthermore, periodicity appears also in (the otherwise quite complex) social networks which describe the dynamics of people meeting~\cite{snapnets,sapiezynski2015tracking}, as every person individually follows mostly a weekly routine.%~\cite{Arrighi2023Multi}.

Expanding the work on periodic temporal graphs have already been studied 
(see~\cite[Class 8]{casteigts2012time} and~\cite{Arrighi2023Multi,ErlebachS20,morawietz2021timecop,morawietz2020timecop}), 
our study represents the first attempt to understand the complexity of a graph realization problem in the context of temporal graphs. 
Therefore, we focus in this paper on the most fundamental case, where all edges have the same period $\Delta$ 
(while in the more general case, each edge $e$ in the underlying graph has a period $\Delta_e$).
As it turns out, the periodic temporal graph realization problem with respect to a given $n \times n$ matrix $D$ of the fastest duration times has a very different computational complexity behavior than the classic graph realization problem with respect to shortest path distances in static graphs.

@@

\subparagraph{Further parameterizations.} There are several structural parameters which can be considered to obtain tractability which are either larger than or incomparable to the feedback vertex number.
\begin{itemize}
    \item The \emph{vertex cover number} measures the distance to an independent set, on which we trivially only have no-instances of our problem. We believe this is a promising parameter to obtain tractability.
    \item The \emph{tree-depth} measures ``star-likeness'' of a graph and is incomparable to both the feedback vertex number and the feedback edge number. We leave the parameterized complexity of our problem with respect to this parameter open.
    \item Parameters that measure ``path-likeness'' such as the \emph{pathwidth} or the \emph{vertex deletion distance to disjoint paths} are also natural candidates to investigate.
\end{itemize}
Furthermore, we can consider combining a structural parameter with $\Delta$. Our NP-hardness reduction (\cref{thm:NPhardness}) produces instances with constant $\Delta$, so as a single parameter $\Delta$ cannot yield fixed-parameter tractability. However, in our parameterized hardness reduction (\cref{thm:W1wrtFVS}) the value for $\Delta$ in the produced instance is large. This implies that our result does not rule out e.g.\ fixed-parameter tractability for the combination of the treewidth and $\Delta$ as a parameter. We believe that investigating such parameter combinations is a promising future research direction.

\subparagraph{Further problem variants.}
There are many natural variants of our problem that are well-motivated and warrant consideration. In the following, we give two specific examples. We believe that one of the most natural generalizations of our problem is to allow more than one label per edge in every $\Delta$-period. A well-motivated variant (especially from the network design perspective) of our problem would be to consider the entries of the duration matrix $D$ as upper-bounds on the duration of fastest paths rather than exact durations. 
%Our work gives a starting point for many interesting future research directions such as the two mentioned examples.
